Computing solution space properties by generic programming tensor networks \thanks{\funding{...}}

### 2022-04-01 paper/paper.tex
Updates from Overleaf

For a general graph, it is computationally inefficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products. %, which scales exponentially as the number of vertices.
%\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a randomly structured tensor network is: first find a good pair-wise tensor contraction order as a binary tree, and then contract two tensors a time by this order.
A contraction order of a tensor network corresponds to a tree decomposition of the line graph of its hypergraph representation and the largest intermediate tensor has a rank equal to the width of this tree decomposition~\cite{Markov2008}.
An optimal (in terms of space complexity) contraction order corresponds to the tree decomposition with smallest width, i.e. the treewidth.
Here, this line graph is equivalent to the graph this tensor network mapped from. 
In practice, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth itself is NP-hard.
However, it is easy to find a close to optimal contraction order within typically a few minutes with a heuristic algorithm~\cite{Kourtis2019,Kalachev2021}.
%The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
For large scale applications, it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
loop and accumulate over certain degrees of freedom so that one can have a smaller tensor network inside the loop due to the removal of these degrees of freedom.

@@

\centerline{\begin{tikzpicture}[]
    \def\dx{0};
    \def\dy{0};
    \def\d{1};
    \def\a{1.0};
    \def\b{1};
    \def\ya{-1.0};
    \def\yb{-0.8};
    \node[] at (\dx+5*\a, \dy+5*\b) (R1) {$R[n^2]$};
    \node[] at (\dx+3*\a, \dy+4*\b) (S1) {$S_{bc}[n^3]$};
    \node[] at (\dx+2*\a, \dy+3*\b) (Q1) {$Q_{bd}[n^2]$};
    \node[] at (\dx+8.5*\a, \dy+3*\b) (Q2) {$Q_{bc}[n^2]$};
    \node[] at (\dx+\a, \dy+2*\b) (P1) {$P_{d}[n]$};
    \node[] at (\dx+9*\a, \dy+2*\b) (P2) {$P_{bc}[n^3]$};
    \node[] at (\dx+0.5*\d, \dy+\b) (M1) {$M_{d}[n^2]$};
    \node[] at (\dx+3.5*\d, \dy+\b) (M2) {$M_{bd}[n^2]$};
    \node[] at (\dx+5.5*\d, \dy+\b) (M3) {$M_{cd}[n^2]$};
    \node[] at (\dx+9.5*\d, \dy+\b) (M4) {$M_{ac}[n^2]$};
    \node[] at (\dx, \dy) (Bde) {$B_{de}$};
    \node[] at (\dx+\d, \dy) (We) {$W_e$};
    \node[] at (\dx+2*\d, \dy) (Wd) {$W_d$};
    \node[] at (\dx+3*\d, \dy) (Bbd) {$B_{bd}$};
    \node[] at (\dx+4*\d, \dy) (Wb) {$W_b$};
    \node[] at (\dx+5*\d, \dy) (Bcd) {$B_{cd}$};
    \node[] at (\dx+6*\d, \dy) (Wc) {$W_c$};
    \node[] at (\dx+7*\d, \dy) (Bbc) {$B_{bc}$};
    \node[] at (\dx+8*\d, \dy) (Bab) {$B_{ab}$};
    \node[] at (\dx+9*\d, \dy) (Bac) {$B_{ac}$};
    \node[] at (\dx+10*\d, \dy) (Wa) {$W_a$};
    \draw[] (Bde) -- (M1);
    \draw[] (We) -- (M1);
    \draw[] (Bbd) -- (M2);
    \draw[] (Wb) -- (M2);
    \draw[] (Bcd) -- (M3);
    \draw[] (Wc) -- (M3);
    \draw[] (Bac) -- (M4);
    \draw[] (Wa) -- (M4);
    \draw[] (M1) -- (P1);
    \draw[] (Wd) -- (P1);
    \draw[] (Bab) -- (P2);
    \draw[] (M4) -- (P2);
    \draw[] (P1) -- (Q1);
    \draw[] (M2) -- (Q1);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Q1) -- (S1);
    \draw[] (M3) -- (S1);
    \draw[] (S1) -- (R1);
    \draw[] (Q2) -- (R1);
\end{tikzpicture}}
% \small{
% \begin{align*}
%     &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a, w_a)_{s_a} W(x_b, w_b)_{s_b} W(x_c, w_c)_{s_c} W(x_d, w_d)_{s_d} W(x_e, w_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
%     =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e,w_e)_{s_e}\right) W(x_d,w_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b, w_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c, w_c)_{s_c}\right)\right)\right.\\
%     &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a,w_a)_{s_a}\right)\right)\right)\right)\\
%     =&1 + x_a^{w_a} + x_b^{w_b} + x_c^{w_c} + x_d^{w_d} + x_e^{w_e} + x_a^{w_a}x_d^{w_d} + x_a^{w_a}x_e^{w_e} + x_c^{w_c}x_e^{w_e} + x_b^{w_b}x_e^{w_e}
%     %=&1+5x+4x^2 \qquad \quad (x_{i} = x, w_{i}=1).
% \end{align*}}
The contraction process goes from bottom to top, where the root node stores the contraction result, the leaves are input tensors and the rest nodes are all intermediate contraction results.
Tensor subscripts are indices so that the number of subscripts indicates the space complexity to store this tensor. The contraction complexity to generate a tensor is annotated in the square brackets, where $n$ is the dimension of degrees of freedom, which in our case is $2$.
One can easily check the largest tensor during computing has space complexity $O(n^2)$ and this is optimal, i.e. the treewidth of the original $5$-vertex graph is $2$. The time complexity is $O(n^3)$, which is much smaller than that of direct enumeration $O(n^5)$.
\end{example}

@@

By assigning a real number to $x$, one can evaluate this independence polynomial for this specific value directly by contracting this tensor network, e.g. the total number of independent sets is equal to $I(G, 1)$.
However, instead of evaluating this polynomial for a certain value, we are more interested in knowing the coefficients of this polynomial, because this quantity tells us the counting of independent sets at each size.
To achieve this, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x)$ and $B$ from the standard number types to the additive identity,
$\mymathbb{0}$, and multiplicative identity, $\mymathbb{1}$, of a commutative semiring as discussed in Sec.~\ref{sec:generic}.

@@

%These operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring. The tensors $W$ and $B$ can thus be written as 

@@

For practical methods to evaluate these coefficients, we refer readers to \App{sec:finitefield}, where we provide an accurate and memory efficient method to find the polynomial by contracting and fitting with finite field algebra.

>>>>

For a general graph, it is computationally inefficient to evaluate \Eq{eq:idp} by directly summing up the $2^{|V|}$ products. %, which scales exponentially as the number of vertices.
%\purple{Perhaps explicitly say what this scales as?}\blue{I am not sure explicitly saying ``scales exponentially as the number of vertices'' is helpful, because treewidth often scales linearly as number of vertices, so tensor network also scales exponentially as the number vertices in most cases.}
The standard approach to evaluate a randomly structured tensor network is: find a good pair-wise tensor contraction order as a binary tree, and then contract two tensors a time by this order.
Let us denote the line graph of the hypergraph representation of a tensor network as $G'$.
A contraction order of this tensor network corresponds to a tree decomposition of $G'$ and the largest intermediate tensor has a rank equal to the width of this tree decomposition~\cite{Markov2008}.
Therefore, an optimal (in terms of space complexity) contraction order corresponds to the tree decomposition of $G'$ with smallest width (or the treewidth ${\rm tw}(G')$).
In the independent set problem, treewidth of $G'$ is the same to the treewidth of the graph this tensor network mapped from. 
In practice, it is difficult to find an optimal contraction order for large tensor networks because finding the treewidth is well known NP-hard problem.
However, it is easy to find a close to optimal contraction order within typically a few minutes with a heuristic algorithm~\cite{Kourtis2019,Kalachev2021}.
%The pairwise tensor contraction also makes it possible to utilize basic linear algebra subprograms (BLAS) functions to speed up the computation for certain tensor element types.
For large scale applications, it is also possible to slice over certain degrees of freedom to reduce the space complexity, i.e.
loop and accumulate over certain degrees of freedom so that one can have a smaller tensor network inside the loop due to the removal of these degrees of freedom.

@@

\centerline{\begin{tikzpicture}[]
    \def\dx{0};
    \def\dy{0};
    \def\d{1};
    \def\a{1.0};
    \def\b{1};
    \def\ya{-1.0};
    \def\yb{-0.8};
    \node[] at (\dx+5*\a, \dy+5*\b) (R1) {$R[n^2]$};
    \node[] at (\dx+3*\a, \dy+4*\b) (S1) {$S_{bc}[n^3]$};
    \node[] at (\dx+2*\a, \dy+3*\b) (Q1) {$Q_{bd}[n^2]$};
    \node[] at (\dx+8.5*\a, \dy+3*\b) (Q2) {$Q_{bc}[n^2]$};
    \node[] at (\dx+\a, \dy+2*\b) (P1) {$P_{d}[n]$};
    \node[] at (\dx+9*\a, \dy+2*\b) (P2) {$P_{bc}[n^3]$};
    \node[] at (\dx+0.5*\d, \dy+\b) (M1) {$M_{d}[n^2]$};
    \node[] at (\dx+3.5*\d, \dy+\b) (M2) {$M_{bd}[n^2]$};
    \node[] at (\dx+5.5*\d, \dy+\b) (M3) {$M_{cd}[n^2]$};
    \node[] at (\dx+9.5*\d, \dy+\b) (M4) {$M_{ac}[n^2]$};
    \node[] at (\dx, \dy) (Bde) {$B_{de}$};
    \node[] at (\dx+\d, \dy) (We) {$W_e$};
    \node[] at (\dx+2*\d, \dy) (Wd) {$W_d$};
    \node[] at (\dx+3*\d, \dy) (Bbd) {$B_{bd}$};
    \node[] at (\dx+4*\d, \dy) (Wb) {$W_b$};
    \node[] at (\dx+5*\d, \dy) (Bcd) {$B_{cd}$};
    \node[] at (\dx+6*\d, \dy) (Wc) {$W_c$};
    \node[] at (\dx+7*\d, \dy) (Bbc) {$B_{bc}$};
    \node[] at (\dx+8*\d, \dy) (Bab) {$B_{ab}$};
    \node[] at (\dx+9*\d, \dy) (Bac) {$B_{ac}$};
    \node[] at (\dx+10*\d, \dy) (Wa) {$W_a$};
    \draw[] (Bde) -- (M1);
    \draw[] (We) -- (M1);
    \draw[] (Bbd) -- (M2);
    \draw[] (Wb) -- (M2);
    \draw[] (Bcd) -- (M3);
    \draw[] (Wc) -- (M3);
    \draw[] (Bac) -- (M4);
    \draw[] (Wa) -- (M4);
    \draw[] (M1) -- (P1);
    \draw[] (Wd) -- (P1);
    \draw[] (Bab) -- (P2);
    \draw[] (M4) -- (P2);
    \draw[] (P1) -- (Q1);
    \draw[] (M2) -- (Q1);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Bbc) -- (Q2);
    \draw[] (P2) -- (Q2);
    \draw[] (Q1) -- (S1);
    \draw[] (M3) -- (S1);
    \draw[] (S1) -- (R1);
    \draw[] (Q2) -- (R1);
\end{tikzpicture}}
% \small{
% \begin{align*}
%     &\sum_{s_a,s_b,s_c,s_d,s_e} W(x_a, w_a)_{s_a} W(x_b, w_b)_{s_b} W(x_c, w_c)_{s_c} W(x_d, w_d)_{s_d} W(x_e, w_e)_{s_e} B_{s_a s_b} B_{s_b s_d} B_{s_c s_d} B_{s_a s_c} B_{s_b s_c} B_{s_d s_e}\\
%     =&\sum_{s_b,s_c}\left(\sum_{s_d}\left(\left(\left(\left(\sum_{s_e}B_{s_d s_e}W(x_e,w_e)_{s_e}\right) W(x_d,w_d)_{s_d}\right) \left(B_{s_bs_d} W(x_b, w_b)_{s_b}\right)\right) \left(B_{s_cs_d} W(x_c, w_c)_{s_c}\right)\right)\right.\\
%     &\phantom{XXX}\left.\left(B_{s_bs_c}\left(\sum_{s_a}B_{s_as_b}\left(B_{s_as_c}W(x_a,w_a)_{s_a}\right)\right)\right)\right)\\
%     =&1 + x_a^{w_a} + x_b^{w_b} + x_c^{w_c} + x_d^{w_d} + x_e^{w_e} + x_a^{w_a}x_d^{w_d} + x_a^{w_a}x_e^{w_e} + x_c^{w_c}x_e^{w_e} + x_b^{w_b}x_e^{w_e}
%     %=&1+5x+4x^2 \qquad \quad (x_{i} = x, w_{i}=1).
% \end{align*}}
The contraction process goes from bottom to top, where the root node stores the contraction result; the leaves are input tensors and the rest nodes are all intermediate contraction results.
Tensor subscripts are indices so that the number of subscripts indicates the space complexity to store this tensor.
The contraction complexity to generate a tensor is annotated in the square brackets, where $n$ is the dimension of degree of freedoms, which is $2$ in a tensor network mapped from an independent set problem.
One can easily check the largest tensor in contraction has space complexity $O(n^2)$ and this is the smallest over all possible contraction trees, i.e. the treewidth of the original $5$-vertex graph is $2$. The time complexity is $O(n^3)$, which is much smaller than that of direct enumeration $O(n^5)$.
\end{example}

@@

By assigning a real number to $x$, one can evaluate this independence polynomial for this specific value directly using tensor network contraction.
It is already very useful, for example, the total number of independent sets can be evaluated as $I(G, 1)$.
However, instead of evaluating this polynomial for a certain value, we are more interested in knowing the coefficients of this polynomial, because this quantity tells us the counting of independent sets at different sizes.
To this end, let us first elevate the tensor elements $0$s and $1$s in tensors $W(x)$ and $B$ from the standard number types to the additive identity, $\mymathbb{0}$, and multiplicative identity, $\mymathbb{1}$, of a commutative semiring as discussed in Sec.~\ref{sec:generic}.

@@

%These operations are standard addition and multiplication operations of polynomials, and the polynomial type forms a commutative ring.
The tensors $W$ and $B$ can thus be written as 

@@

For practical methods to evaluate these coefficients, we refer readers to \App{sec:finitefield}, where we provide an accurate and memory efficient method to find the polynomial by contracting and fitting on finite field algebra.
